Statically Typed String Sanitation Inside a Python (Technical Report)

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
Updated TR

\newtheorem{tr}{TR Item}
\renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
\newtheorem{trexam}[tr]{Example}
\newtheorem{trthm}[tr]{Theorem}
\newtheorem{trlem}[tr]{Lemma}
\newtheorem{trdef}[tr]{Definition}
\newtheorem{trconj}[tr]{Conjecture}
\newtheorem{ass}[tr]{Assumption}

@@

\title{Statically Typed String Sanitation Inside a Python (Technical Report)}
\author{Nathan Fulton \and Cyrus Omar \and Jonathan Aldrich}
\date{December 2014}
\trnumber{CMU-ISR-14-112}
\keywords{type systems; regular languages; input sanitation; string sanitation}
\support{This work was supported by the National Security Agency lablet contract \#H98230-14-C-0140.}
\abstract{This report contains supporting evidence for claims put forth and explained in the paper ``Statically Typed String Sanitation Inside a Python'' \cite{fulton2014}, including proofs of lemmas
and theorems asserted in the paper, examples, and additional discussion of the paper's technical content.}
\begin{document}
\maketitle
\section{Terminology and Notation}
Theorems and lemmas appearing in \cite{fulton2014} are numbered, while supporting facts appearing only in the Technical Report are lettered.
Numbered items correspond to the numbering in \cite{fulton2014}.

\section{$\lambdas$}

This section presents proofs of lemmas and theorems about the type systems
presented in \cite{fulton2014}, the accompanying paper. In addition, we provide some examples to help 
motivate and explain definitions.

To facilitate the type safety proof, we introduce
a small step semantics for both $\lambdas$ and $\lambdap$. All theorems in
this section are proven as stated in \cite{fulton2014}.

\subsection{Head and Tail Operations}

\begin{defn}[Definition of \lhead{r}]\label{def:lhead}
The relation $\lhead{r}=r'$ is defined in terms of the structure of $r$:
\begin{align*}
\lhead{r}                 &= \lhead{r, \epsilon} \\
\lhead{\epsilon, r'}      &= \epsilon \\
\lhead{a, r'}             &= a \\
\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2} \\
\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} + \lhead{r_2, r'} \\
\lhead{r^*, r'}           &= \lhead{r', \epsilon} + \lhead{r, \epsilon}
\end{align*}
\end{defn}

@@

\begin{defn}[Definition of \ltail{r}]
The relation $\ltail{r} = r'$ is defined in terms of $\lhead{r}$.
Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r) + \epsilon$.
\end{defn}

Using these definitions of head and tail, we establish a correctness
result upon which type soundness for the concatenation operator in
$\lambdas$ depends.

Throughout the following proofs, we abbreviate $s \in \lang{r}$ by $s \in r$.

\begin{trlem}[Leading characters are in the head] \label{trlemmax}
If $s = c_1  c_2 \cdots c_m \in \lang{r}$, then
$c_1 \in \lhead{r}$.
\end{trlem}
\begin{proof}
By structural induction on $r$.

If $r = x$ for some character $x$ then $s \in \lang{r}$ implies $s = x$.
By definition of lhead, $\lhead{s} = \lhead{s, \epsilon} = \lhead{x, \epsilon} = x$.
Finally, $x \in \lang{x}$.

If $r = r_1 \cdot r_2$ then $s \in \lang{r}$ implies $s = c_1 s_1 s_2$
such that $c_1 s_1 \in \lang{r_1}$. 
By definition of lhead, $\lhead{r_1 \cdot r_2} = \lhead{r_1, r_2} =
\cdots + \lhead{r_1, \epsilon} = \cdots + \lhead{r_1}$. Therefore, $r_1 \subseteq \lhead{r_1, r_2}$
and so it suffices to show that $s = c_1 s_1 s_2$ such that $c_1 \in \lhead{r_1}$,
\todo{this is why we break out the $c_1$ from the $s_1$, both here and going forward}
which holds by induction because $c_1 s_1 \in \lang{r_1}$.

If $r = r_1 + r_2$ then $s \in r$ implies either $s \in r_1$ or $s \in r_2$.
Without loss of generality suppose $s \in r_1$.
Then by induction $s = c_1 \cdots c_m$ such that $c_1 \in \lhead{r_1}$.
By definition, $\lhead{r} = \lhead{r_1, \epsilon} + \lhead{r_2, \epsilon} = \lhead{r_1} + \cdots$.
Therefore, $c_1 \in \lhead{r}$.

If $r = r_1^*$ then $s \in r$ implies that $s = s_1 \cdots s_k$ where $k \ge 0$ and $s_i \in \lang{r}$.
By definition of lhead, $\lhead{r_1^*} = \epsilon + \lhead{r_1}$.
If $s = \epsilon$, then $s \in \lang{\epsilon} \subseteq \lang{\lhead{r_1}}$.
If $s = s_1 \cdots s_k$, then $s_1 \in \lang{r}$ and so by induction $s_1 = c_1 \cdots c_k$ such that $c_1 \in \lang{\lhead{r}}$.
\end{proof}

\begin{trthm}[Correctness of \textsf{ltail}]
If $s \in \lang{r}$ then $s \in \lang{\lhead{r} \cdot \ltail{r}}$.
\end{trthm}
\begin{proof}
The proof proceeds by structural induction on $r$.
In each case, we identify sublanguages of the head and tail of $r$ such that 
$s$ is in the concatenation of the the sublanguages.
Throughout the proof, we use the fact that $\lang{r'} \subseteq \lang{r}$ and $\lang{q'} \subseteq \lang{q}$
implies $\lang{r' \cdot q'} \subseteq \lang{r \cdot q}$.

\item \textbf{Case $r = c$ for $c \in Sigma$}.
By definition, if $s \in \lang{c}$ then $s = c$.
By definition of lhead and ltail, $\lhead{c} = \lhead{c, \epsilon} = c$
and $\lang{\epsilon} \subseteq \lang{\ltail{r}}$.
Therefore, $c \in \lang{c \cdot \epsilon} \subseteq \lang{\lhead{r} + \ltail{r}}$.

\item \textbf{Case $r = r_1 \cdot r_2$}.
Suppose $s \in \lang{r}$.
We may decompose $s$ as $s_1  \cdots s_n$ such that $s_1 = c_1 s_1' \in \lang{r_1}$.
Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
Therefore, $c_1 \in \lhead{r_1 \cdot r_2}$ by definition of lhead.
It suffices to show that $s_1' s_2 \in \lang{\delta_{c_1}(r)}$.
Since $s = c_1 s_1' s_2\in \lang{r}$, it follows by the definition of derivative that $s_1' s_2 \in \delta_{c_1}(r)$.
Therefore, $s = c_1 s_1' s_2 \in \lang{ c_1 \cdot \delta_{c_1} } \subseteq \lang{ \lhead{r} \cdot \ltail{r}}$.

\item \textbf{Case $r = r_1 + r_2$}.
Suppose $s \in \lang{r_1 + r_2}$
so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
Consider without loss of generality the case where $s \in \lang{r_1}$.
By induction, $s \in \lang{\lhead{r_1} \cdot \ltail{r_1}}$
Note that $\lhead{r_1} \subseteq \lhead{r}$
and $\ltail{r_1} \subseteq \ltail{r}$.
Therefore, $\lhead{r_1} \cdot \ltail{r_2} \subseteq \lhead{r_1} \cdot \ltail{r_1}$.

\item \textbf{Case $r = q^*$}.
Either $s$ is the empty string or else the $k^{th}$ unwinding (i.e., $s = s_1 \cdots s_k$ where $k > 0$ and $s_i \in \lang{q}$).

Suppose $s = \epsilon$.
By definition of lhead, $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + \cdots = \epsilon + \cdots$.
By definition of ltail, $\ltail{r} = \epsilon \cdots$.
From these facts we conclude that $s = \epsilon \in \lang{\epsilon \cdot \epsilon} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$.

Suppose instead that $s = s_1 \cdots s_k$ for $k < 0$ where $s_i \in \lang{q}$.
If $q = \epsilon$ then $s = \epsilon$ and the conclusion follows for the same reason as the epsilon subcase of the character case.
Otherwise, $s$ is not the empty string and so $s = c_1 s_1 s_2$ where $s_1$ and $s_2$ might be empty.
By the lemma above, $c_1 \in \lhead{s}$.
By the definition of derivative, $s_1s_2 \in \delta_{c_1}(s)$.
Therefore, $s \in \lang{\lhead{s} \cdot \delta_{c_1}(s)} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$, where the set inclusion follows by the definition
of lhead and ltail.
\end{itemize}
\end{proof}

\begin{trexam}[Decomposition is not Reversible]\label{exam:headsandtails}
$r \not = \lhead{r} \cdot \ltail{r}$.
\end{trexam}
\begin{proof}
A simple counter-example is $ab + cd$. 
Note that $\lhead{a\cdot b + c\cdot d} = a+c$ and $\ltail{a\cdot b + c\cdot d} = b + d$. 
Therefore, $\{ad,bc\} \subset \lang{\lhead{a\cdot b + cd\cdot } \cdot \ltail{a\cdot b + c\cdot d}}$ even though neither of these is in $\lang{r}$.
\end{proof}

Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
property required for soundness.

This is a general pattern in string operations: $\lambdas$ simulates -- within the type system -- common operations on strings. 
If there is an operation for concatenating to strings, we define an operation for concatenating two regular expressions.
If there is an operation for peeling off the first (n) characters of a string, then we define an operation for 
converting a regular expression $r$ into a regular expression $r'$ which recognizes any $n^{th}$ suffix of a string in $r$.

It is important to note, however, that the type system need not \emph{exactly} simulate the action of string operations.
In the case of concatenation, we lose some information because more string values are possible -- according the types --
than are actually possible in the dynamic semantics.
Soundness is not lost because the types are conservative in their approximation.

In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this technical report.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Substitution.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Some Corollaries About Substitution and Language Replacement}

\begin{defn}[$\tt{subst}$]
We consider several choices in the string operations section.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
We consider several choices in the string operations section.
\end{defn}

\begin{prop}[Closure] \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{prop}
\begin{proof}
This result is proven for various formulations in the next section.
\end{proof}

\begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{prop}
\begin{proof}
This is exactly the correctness result proven for some pairs of ${\tt subst}$ and ${\tt replace}$ in the previous section.
\end{proof}

\begin{ass}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{enumerate}

Lemma~\ref{thm:regexprops} states some well-known properties about regular expressions.

%\begin{lem}
%  If $\emptyset \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
%\end{lem}
%\begin{proof}
%  The proof proceeds by induction on the derivation of the premise.
%  The only non-trivial cases (those which require more than an appeal to inversion) are S-T-Case, S-T-Replace and S-T-Concat. 
%
%In the S-T-Case case, note that ${\sf lhead}$ and ${\sf ltail}$ are total functions for well-formed regular expressions to well-formed regular expressions.
%
%In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.
%
%In the S-T-Replace case, it suffices to show that $\lreplace{r}{r_1}{r_2}$ is a regular expression assuming (inductively) that $r, r_1$ and $r_2$ are all regular expressions.
%This follows from the Closure proposition.
%\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Type Safety and Security.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The Small Step Semantics}

To prove type safety and the security theorems for the big step semantics, we
first prove type safety for a small step semantics in Figure~\ref{fig:sdlambdas}
and then extend this to the big step semantics in Figure~\ref{fig:dlambdas} by
proving a correspondence between the semantics.

@@

\begin{trlem}[Canonical Forms]\label{cfgen}
If $\emptyset \vdash v : \sigma$ then:
\begin{enumerate}
\item If $\sigma = \stringin{r}$ then $u = \rstr{s}$.
\item If $\sigma = \sigma_1 \rightarrow \sigma_2$ then $v = \lambda x . e'$ and $s \in \lang{r}$.
\end{enumerate}

\end{trlem}
\begin{proof}
By inspection of valuation and typing rules.
\end{proof}

For the sake of completeness, we include a statement of the weaker lemma
stated in the paper:

\begin{lem}[Canonical Forms for the String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lem}
\begin{proof}
This fact follows directly from Lemma~\ref{cfgen} and the preservation lemma.
\end{proof}

\begin{trlem}[Progress of small step semantics.]
If $\emptyset\vdash e : \sigma$ either $e = v$ or $e \mapsto e'$ for some $e'$.
\end{trlem}
\begin{proof}
The proof proceeds by induction on the derivation of $\emptyset\vdash e : \sigma$.
\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of progress for
the simply typed lambda calculus.

\item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
Then $\rstr{s} \ \val$ by SS-E-RStr.

\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{r}$ and $\emptyset e_1 : \stringin{r_1}$ and $\emptyset \vdash e_2 : \stringin{r_2}$.
By induction, $e_1 \mapsto e_1'$ or $e_1 = v_1$ and similarly
for $e_2$. 
If $e_1$ steps, then SS-E-Concat-Left applies and so
$\rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$. Similarly, if $e_2$ steps
then $e$ steps by SS-E-Concat-Right.

@@

\def\mline#1{\par\hspace*{-\leftmargin}\parbox{\textwidth}{\[#1\]}}
\item \textbf{S-T-Case}.
Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$ and $\emptyset\vdash e_1 : \stringin{r}$.
%\begin{align}
%&\vdash e_1 : \stringin{r} \\
%&\vdash e_2 : \sigma \\
%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma
%\end{align}
%Actually this isn't even necessary...
%By (2), $e_2 \mapsto e_2'$ or $e_2 \val$. 
%In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
By induction and Lemma~\ref{thm:cfs} it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
In the former case, $e$ steps by S-E-Case-Left.
In the latter case, note that $s = \epsilon$ or $s = at$ for some string $t$.
If $s = \epsilon$ then $e$ steps by S-E-Case-$\epsilon$-Val,
and if $s = at$ the $e$ steps by S-E-Case-Concat.

>>>>

\newtheorem{tr}{TR Item}
\renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
\newtheorem{trexam}[tr]{Example}
\newtheorem{trthm}[tr]{Theorem}
\newtheorem{trlem}[tr]{Lemma}
\newtheorem{trdef}[tr]{Definition}
\newtheorem{trconj}[tr]{Conjecture}
\newtheorem{ass}[tr]{Assumption}
\newtheorem{condition}[tr]{Condition}

@@

\title{Statically Typed String Sanitation Inside a Python (Technical Report)}
\author{Nathan Fulton \and Cyrus Omar \and Jonathan Aldrich}
\date{December 2014}
\trnumber{CMU-ISR-14-112}
\keywords{type systems; regular languages; input sanitation; string sanitation}
\support{This work was supported by the National Security Agency lablet contract \#H98230-14-C-0140.}
\abstract{This report contains supporting evidence for claims put forth and explained in the paper ``Statically Typed String Sanitation Inside a Python'' \cite{fulton2014}, including proofs of lemmas
and theorems asserted in the paper, examples, additional discussion of the paper's technical content, and errata.}
\begin{document}
\maketitle

\tableofcontents
\listoffigures
\clearpage

\section{Terminology and Notation}
Theorems and lemmas appearing in \cite{fulton2014} are numbered correspondingly, while supporting facts appearing only in the Technical Report are lettered.

\section{Regular Expressions}
The syntax of regular expressions over some alphabet $\Sigma$ is shown in Figure \ref{fig:regex}.

\begin{ass}[Regular Expression Congruences]
We assume regular expressions are implicitly identified up to the following congruences:
\begin{align*}
\epsilon\cdot r & \equiv r\\
r \cdot \epsilon & \equiv r\\
(r_1\cdot r_2) \cdot r_3 & \equiv r_1 \cdot (r_2 \cdot r_3)\\
r_1 + r_2 &\equiv r_2 + r_1\\
(r_1 + r_2) + r_3 &\equiv r_1 + (r_2 + r_3)\\
\epsilon^* & \equiv \epsilon
\end{align*}
\end{ass}

\begin{ass}[Properties of Regular Languages] \label{thm:regexprops}
We assume the following properties:
\begin{enumerate}

\section{$\lambdas$}
The syntax of $\lambdas$ is specified in Figure \ref{fig:glambdas}. The static semantics is specified in Figure \ref{fig:slambdas}.

\subsection{Head and Tail Operations}
The following correctness conditions must hold for any definition of $\lhead{r}$ and $\ltail{r}$.

\begin{condition}[Correctness of Head] \label{trlemmax}
If $c_1  s' \in \lang{r}$, then
$c_1 \in \lang{\lhead{r}}$.
\end{condition}

\begin{condition}[Correctness of Tail] \label{trlemma:tail-correctness}
If $c_1 s' \in \lang{r}$ then $s' \in \lang{\ltail{r}}$.
\end{condition}

For example, we conjecture (but do not here prove) that the definitions below satisfy these conditions. Note that these are slightly amended relative to the published paper.

\begin{defn}[Definition of $\lhead{r}$]\label{def:lhead}
We first define an auxiliary relation that determines the set of characters that the head might be, tracking the remainder of any sequences that appear:
\begin{align*}
\lhead{\epsilon, \epsilon} &= \emptyset\\
\lhead{\epsilon, r'}      &= \lhead{r', \epsilon} \\
\lhead{a, r'}             &= \{a\} \\
\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2 \cdot r'} \\
\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} \cup \lhead{r_2, r'} \\
\lhead{r^*, r'}           &= \lhead{r, \epsilon} \cup \lhead{r', \epsilon}
\end{align*}
We define $\lhead{r} = a_1 + a_2 + ... + a_i$ iff $\lhead{r, \epsilon} = \{a_1, a_2, ..., a_i\}$.
\end{defn}

% \begin{trlem}[Correctness of Head] \label{trlemmax}
% If $c_1  s' \in \lang{r}$, then
% $c_1 \in {\lhead{r,\epsilon}}$.
% \end{trlem}
% \begin{proof}
% By structural induction on $r$.

% The case for $r = \epsilon$ does not apply because $s$ is non-empty.

% If $r = a$ for some character $a$ then $s \in \lang{r}$ implies $c_1 = a$.
% By definition $\lhead{a, \epsilon} = \{a\}$.
% Finally, $a \in \{a\}$ by definition.

% If $r = r_1 \cdot r_2$ then we proceed by case analysis on $r_1$. 
% \begin{quote}
% We need not consider the case where $r_1=\epsilon$ by congruence.

% If $r_1=a$ then $c_1=a$ and $\lhead{a \cdot r_2, \epsilon}=\lhead{a, r_2}=a$ and $a \in \{a\}$ by definition.

% If $r_1=r_{11}\cdot r_{12}$, then we have that $\lhead{r_{11} \cdot r_{12} \cdot r_2, \epsilon}=\lhead{r_{11}, r_{12}\cdot r_2}$. By the IH we have that $c_1 \in {\lhead{r_{11}\cdot r_{12}, \epsilon}}$. By Lemma \ref{trlemma:head-concatenation-1}, we have that $c_1 \in \lhead{r_{11}\cdot r_{12} \cdot r_2, \epsilon}$.

% If $r_1=r_{11} + r_{12}$, then by definition we have that

% \end{quote}

% If $r = r_1 \cdot r_2$ then $s \in \lang{r}$ implies $s = c_1 s_1 s_2$
% such that $c_1 s_1 \in \lang{r_1}$. 
% By definition of lhead, $\lhead{r_1 \cdot r_2} = \lhead{r_1, r_2} =
% \cdots + \lhead{r_1, \epsilon} = \cdots + \lhead{r_1}$. Therefore, $r_1 \subseteq \lhead{r_1, r_2}$
% and so it suffices to show that $s = c_1 s_1 s_2$ such that $c_1 \in \lhead{r_1}$,
% \todo{this is why we break out the $c_1$ from the $s_1$, both here and going forward}
% which holds by induction because $c_1 s_1 \in \lang{r_1}$.

% If $r = r_1 + r_2$ then $s \in r$ implies either $s \in r_1$ or $s \in r_2$.
% Without loss of generality suppose $s \in r_1$.
% Then by induction $s = c_1 \cdots c_m$ such that $c_1 \in \lhead{r_1}$.
% By definition, $\lhead{r} = \lhead{r_1, \epsilon} + \lhead{r_2, \epsilon} = \lhead{r_1} + \cdots$.
% Therefore, $c_1 \in \lhead{r}$.

% If $r = r_1^*$ then $s \in r$ implies that $s = s_1 \cdots s_k$ where $k \ge 0$ and $s_i \in \lang{r}$.
% By definition of lhead, $\lhead{r_1^*} = \epsilon + \lhead{r_1}$.
% If $s = \epsilon$, then $s \in \lang{\epsilon} \subseteq \lang{\lhead{r_1}}$.
% If $s = s_1 \cdots s_k$, then $s_1 \in \lang{r}$ and so by induction $s_1 = c_1 \cdots c_k$ such that $c_1 \in \lang{\lhead{r}}$.
% \end{proof}

@@

\begin{defn}[Definition of $\ltail{r}$]
If $\lhead{r, \epsilon} = \{a_1, a_2, ..., a_i\}$, then  
we define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
\end{defn}

% Using these definitions of head and tail, we establish a correctness
% result upon which type soundness for the concatenation operator in
% $\lambdas$ depends.

% \begin{trthm}[Correctness of \textsf{ltail}]
% If $s \in \lang{r}$ then $s \in \lang{\lhead{r} \cdot \ltail{r}}$.
% \end{trthm}
% \begin{proof}
% The proof proceeds by structural induction on $r$.
% In each case, we identify sublanguages of the head and tail of $r$ such that 
% $s$ is in the concatenation of the the sublanguages.
% Throughout the proof, we use the fact that $\lang{r'} \subseteq \lang{r}$ and $\lang{q'} \subseteq \lang{q}$
% implies $\lang{r' \cdot q'} \subseteq \lang{r \cdot q}$.

% \begin{itemize}[label=$ $,itemsep=1ex]

% \item \textbf{Case $r = c$ for $c \in Sigma$}.
% By definition, if $s \in \lang{c}$ then $s = c$.
% By definition of lhead and ltail, $\lhead{c} = \lhead{c, \epsilon} = c$
% and $\lang{\epsilon} \subseteq \lang{\ltail{r}}$.
% Therefore, $c \in \lang{c \cdot \epsilon} \subseteq \lang{\lhead{r} + \ltail{r}}$.

% \item \textbf{Case $r = r_1 \cdot r_2$}.
% Suppose $s \in \lang{r}$.
% We may decompose $s$ as $s_1  \cdots s_n$ such that $s_1 = c_1 s_1' \in \lang{r_1}$.
% Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
% Therefore, $c_1 \in \lhead{r_1 \cdot r_2}$ by definition of lhead.
% It suffices to show that $s_1' s_2 \in \lang{\delta_{c_1}(r)}$.
% Since $s = c_1 s_1' s_2\in \lang{r}$, it follows by the definition of derivative that $s_1' s_2 \in \delta_{c_1}(r)$.
% Therefore, $s = c_1 s_1' s_2 \in \lang{ c_1 \cdot \delta_{c_1} } \subseteq \lang{ \lhead{r} \cdot \ltail{r}}$.

% \item \textbf{Case $r = r_1 + r_2$}.
% Suppose $s \in \lang{r_1 + r_2}$
% so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
% Consider without loss of generality the case where $s \in \lang{r_1}$.
% By induction, $s \in \lang{\lhead{r_1} \cdot \ltail{r_1}}$
% Note that $\lhead{r_1} \subseteq \lhead{r}$
% and $\ltail{r_1} \subseteq \ltail{r}$.
% Therefore, $\lhead{r_1} \cdot \ltail{r_2} \subseteq \lhead{r_1} \cdot \ltail{r_1}$.

% \item \textbf{Case $r = q^*$}.
% Either $s$ is the empty string or else the $k^{th}$ unwinding (i.e., $s = s_1 \cdots s_k$ where $k > 0$ and $s_i \in \lang{q}$).

% Suppose $s = \epsilon$.
% By definition of lhead, $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + \cdots = \epsilon + \cdots$.
% By definition of ltail, $\ltail{r} = \epsilon \cdots$.
% From these facts we conclude that $s = \epsilon \in \lang{\epsilon \cdot \epsilon} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$.

% Suppose instead that $s = s_1 \cdots s_k$ for $k < 0$ where $s_i \in \lang{q}$.
% If $q = \epsilon$ then $s = \epsilon$ and the conclusion follows for the same reason as the epsilon subcase of the character case.
% Otherwise, $s$ is not the empty string and so $s = c_1 s_1 s_2$ where $s_1$ and $s_2$ might be empty.
% By the lemma above, $c_1 \in \lhead{s}$.
% By the definition of derivative, $s_1s_2 \in \delta_{c_1}(s)$.
% Therefore, $s \in \lang{\lhead{s} \cdot \delta_{c_1}(s)} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$, where the set inclusion follows by the definition
% of lhead and ltail.
% \end{itemize}
% \end{proof}

% \begin{trexam}[Decomposition is not Reversible]\label{exam:headsandtails}
% $r \not = \lhead{r} \cdot \ltail{r}$.
% \end{trexam}
% \begin{proof}
% A simple counter-example is $ab + cd$. 
% Note that $\lhead{a\cdot b + c\cdot d} = a+c$ and $\ltail{a\cdot b + c\cdot d} = b + d$. 
% Therefore, $\{ad,bc\} \subset \lang{\lhead{a\cdot b + cd\cdot } \cdot \ltail{a\cdot b + c\cdot d}}$ even though neither of these is in $\lang{r}$.
% \end{proof}

% Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
% property required for soundness.

% This is a general pattern in string operations: $\lambdas$ simulates -- within the type system -- common operations on strings. 
% If there is an operation for concatenating to strings, we define an operation for concatenating two regular expressions.
% If there is an operation for peeling off the first (n) characters of a string, then we define an operation for 
% converting a regular expression $r$ into a regular expression $r'$ which recognizes any $n^{th}$ suffix of a string in $r$.

% It is important to note, however, that the type system need not \emph{exactly} simulate the action of string operations.
% In the case of concatenation, we lose some information because more string values are possible -- according the types --
% than are actually possible in the dynamic semantics.
% Soundness is not lost because the types are conservative in their approximation.

% In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
% the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this technical report.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % Substitution.
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Replacement}
The following correctness condition must hold for any definition of $\lreplace{r}{r_1}{r_2}$.

\begin{condition}[Replacement Correctness] \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $$\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$$
\end{condition}

We do not give a particular definition for $\lreplace{r}{r_1}{r_2}$ here.
%\begin{lem}
%  If $\emptyset \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
%\end{lem}
%\begin{proof}
%  The proof proceeds by induction on the derivation of the premise.
%  The only non-trivial cases (those which require more than an appeal to inversion) are S-T-Case, S-T-Replace and S-T-Concat. 
%
%In the S-T-Case case, note that ${\sf lhead}$ and ${\sf ltail}$ are total functions for well-formed regular expressions to well-formed regular expressions.
%
%In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.
%
%In the S-T-Replace case, it suffices to show that $\lreplace{r}{r_1}{r_2}$ is a regular expression assuming (inductively) that $r, r_1$ and $r_2$ are all regular expressions.
%This follows from the Closure proposition.
%\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Type Safety and Security.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Small Step Semantics of $\lambdas$}
Figure~\ref{fig:sdlambdas} specifies a small-step operational semantics for $\lambdas$.

@@

\begin{trlem}[Canonical Forms]\label{cfgen}
If $\emptyset \vdash v : \sigma$ then:
\begin{enumerate}
\item If $\sigma = \stringin{r}$ then $v = \rstr{s}$ and $s \in \lang{r}$.
\item If $\sigma = \sigma_1 \rightarrow \sigma_2$ then $v = \lambda x . e'$.
\end{enumerate}

\end{trlem}
\begin{proof}
By inspection of the static and dynamic semantics.
\end{proof}

\begin{trlem}[Progress]
If $\emptyset\vdash e : \sigma$ either $e = v$ for some $v$ or $e \mapsto e'$ for some $e'$.
\end{trlem}
\begin{proof}
The proof proceeds by rule induction on the derivation of $\emptyset\vdash e : \sigma$.
\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of progress for
the simply typed lambda calculus.

\item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
Then $e=\rstr{s}$.

\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}$ and $\emptyset \vdash e_1 : \stringin{r_1}$ and $\emptyset \vdash e_2 : \stringin{r_2}$.
By induction, $e_1 \mapsto e_1'$ or $e_1 = v_1$ and similarly, $e_2 \mapsto e_2'$ or $e_2 = v_2$. 
If $e_1$ steps, then SS-E-Concat-Left applies and so
$\rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$. Similarly, if $e_2$ steps
then $e$ steps by SS-E-Concat-Right.

@@

\def\mline#1{\par\hspace*{-\leftmargin}\parbox{\textwidth}{\[#1\]}}
\item \textbf{S-T-Case}.
Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$ and $\emptyset\vdash e_1 : \stringin{r}$.
%\begin{align}
%&\vdash e_1 : \stringin{r} \\
%&\vdash e_2 : \sigma \\
%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma
%\end{align}
%Actually this isn't even necessary...
%By (2), $e_2 \mapsto e_2'$ or $e_2 \val$. 
%In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
By induction and Canonical Forms it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
In the former case, $e$ steps by S-E-Case-Left.
In the latter case, note that $s = \epsilon$ or $s = at$ for some string $t$.
If $s = \epsilon$ then $e$ steps by S-E-Case-$\epsilon$-Val,
and if $s = at$ the $e$ steps by S-E-Case-Concat.
